---
title: "The fatal impact of tornadoes and economic effects of floods"
author: "Claus Lv"
date: "Friday, July 25, 2014"
output:
  html_document: default
---

## Synopsis
This report presents an analysis about the impacts of wheather events on the population health and economic. The data is published by the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. We found that the tornado is most harmful for population heath, and the flash flood results in greatest loss on economic value.

## Questions

1. Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?


2. Across the United States, which types of events have the greatest economic consequences?


```{r library}
library(ggplot2)
library(data.table)
```


### Data Processing
Data are downloaded and imported in stormdata data frame
```{r process data}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("tempdata.csv.bz2")){
        download.file(fileUrl, destfile = "tempdata.csv.bz2")
}

stormdata <- read.csv("tempdata.csv.bz2", stringsAsFactors = FALSE)
```

7 variables that are relevante for addressing the questions:

- EVTYPE: Name (category) of the weather event.
- FATALITIES: number of cases with harm to human health
- INJURIES: number of cases with harm to human health
- PROPDMG: Coefficient of amount of property damage (USD)
- PROPDMGEXP: magnitude of property damage (thousands (“K”), millions (“M”)
- CROPDMG: Coefficient of amount of crop damage (USD)
- CROPDMGEXP: magnitude of crop damage 

These are extracted for further analysis.
```{r}
attributes <- c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", 
    "CROPDMG", "CROPDMGEXP")

stormdata <- stormdata[, attributes]
colnames(stormdata) <- c("state", "evtype", "fatalites", "injures", "propdmg", "propdmgexp", 
    "cropdmg", "cropdmgexp")

str(stormdata)
```

The documentation points out that similar storm events can be listed using different wording e.g. “coastal flood” and “coastal flooding.” This is confirmed by a cursory analysis of EVTYPE levels.
```{r}
head(levels(as.factor(stormdata$evtype)),20)
```


### Data Cleaning
Create a function to cleanup the eventType
```{r cleanup eventType}
deliverEvtUnifiedName <- function(dd) {
    evtMatcher <- data.frame(reg = c("NADO|FUNNEL|WATERSPOUT", "THUNDER|STORM|WIND", 
        "HAIL", "FROST|FREEZ|BLIZZARD|WINTER|COLD|LOW TEMP|RECORD LOW|SNOW|ICE", 
        "HEAT|WARM|RECORD HIGH", "COSTAL STORM", "SUNAMI", "RIP CURRENT", "FLASH FLOOD|FLD|FLOOD", 
        "RIVER FLOOD|URBAN FLOOD", "TROPICAL STORM|TROPICAL", "HURRICANE", "DROUGHT", 
        "DUST STORM", "DUST DEVIL", "RAIN", "LIGHTNING"))
    
    factorId <- c("Tornado", "Thunderstorm wind", "Hail", "Cold", "Heat", "Costal Storm", 
        "Sunami", "Rip current", "Flash flood", "River flood", "Tropical Storm", 
        "Hurricane", "Drought", "Dust storm", "Dust devil", "Rain", "Ligntning")
    
    for (i in 1:nrow(evtMatcher)) {
        indexFit <- grep(evtMatcher[i, "reg"], toupper(dd[, "evtype"]))
        if (length(indexFit) > 0) {
            dd[indexFit, "event"] <- factorId[i]
            
        }
    }
    return(dd)
}

```

Cleanup `evtype`
```{r cleanup}
stormdata$event <- ("-")
# Give each event an unified id factor
stormdata <- deliverEvtUnifiedName(stormdata)
otherIndex <- grep("-", stormdata[, "event"])
stormdata[otherIndex, "event"] <- "Other"
stormdata$event <- as.factor(stormdata$event)

# See the event and number of that
table(stormdata$event)

```

Combine coefficients and magnitudes of property and crop damage to form absolute USD amounts and write it back into CROPDMG and PROPDMG collumns.

- Fomula I : propdmgValue = propdmg * propdmgexp

- Fomula II : cropdmgValue = cropdmg * cropdmgexp 


Create function to caculate the property and crop damage by 'dmg' and 'exp' variable
```{r calc damage value}
deliverActualDmgValue <- function(dd) {
    unit <- data.frame(cha = c("B", "M", "K", "H"), val = c(1e+09, 1e+06, 1000, 
        100))
    multi <- c(1e+09, 1e+06, 1000, 100)
    
    for (i in 1:nrow(unit)) {
        # index that match the unit
        indexCrodFit <- grep(unit[i, "cha"], toupper(dd[, "cropdmgexp"]))
        
        if (length(indexCrodFit) > 0) {
            
            # Caculate the actual value
            dd[indexCrodFit, "cropdmgvalue"] <- multi[i] * dd[indexCrodFit, "cropdmg"]
            
        }
        
        indexProdFit <- grep(unit[i, "cha"], toupper(dd[, "propdmgexp"]))
        
        if (length(indexProdFit) > 0) {
            
            dd[indexProdFit, "propdmgvalue"] <- multi[i] * dd[indexProdFit, "propdmg"]
        }
        
    }
    return(dd)
}

stormdata$cropdmgvalue <- stormdata$cropdmg
stormdata$propdmgvalue <- stormdata$propdmg

stormdata <- deliverActualDmgValue(stormdata)


```


Summary Damage Values
```{r}
summary(stormdata$propdmgvalue)
summary(stormdata$cropdmgvalue)

```


Damage on Economic and People:

- Damage on economic : `ecodmgvalue = propdmagevalue + corpdmgvalue `

- Damage on population health : `peodmg = injures + fatalites`

```{r add new variables}
stormdata$ecodmgvalue <- stormdata$cropdmgvalue + stormdata$propdmgvalue
stormdata$peodmg <- stormdata$injures + stormdata$fatalites

```


## Result

Summary Result
```{r Summary}
dt <- data.table(stormdata)
summary <- dt[, list(fatalities = sum(fatalites), injuries = sum(injures), 
                     people_dmg = sum(fatalites, injures), 
                     propdmg = sum(propdmgvalue), cropdmg = sum(cropdmgvalue),
                     eco_dmg = sum(propdmgvalue, cropdmgvalue)), 
                by = event]
head(summary,20)
```


### Question 1
Across the United States, which types of events (use the cleaned `event` variable) are most harmful with respect to population health?

```{r}
summary[order(people_dmg, decreasing = TRUE)][1:5,]

```


Tornados are the most harmful weather events in the United States with respect to population health.

### Question 2
Across the United States, which types of events have the greatest economic consequences?
```{r}
summary[order(eco_dmg, decreasing = TRUE)][1:5,]

```

Flash flood are the most severe events with respect to economic damage.


### Draw figures to solve 2 questions
Data Aggregation
```{r}
library(reshape2)

t <- aggregate(cbind(propdmgvalue, cropdmgvalue, injures, fatalites) ~ event, 
    stormdata, FUN = sum)
tidy <- melt(t, id.var = "event", variable.name = "variable")
colnames(tidy) <- c("event", "damagetype", "value")

head(tidy)

```

### For People Health
```{r}
damgeOnHealth <- tidy[grep("fatalites|injures", tidy[, "damagetype"]), ]

ggplot(damgeOnHealth, 
       aes(x = reorder(event, value), y = value, 
           fill = factor(damagetype, labels = c("Injures", "Fatalites")))) +
        geom_bar(stat = "identity") + 
        labs(title = "Top harmful weather event for populaiton health", 
             x = "Event", y = "Damage on populaiton health (number of people)") + 
        guides(fill = guide_legend(title = "Type of damage")) + 
        xlab("Weather Event") +
        coord_flip()


```


Result: Tornados are the most harmful weather events in the United States with respect to population health.

### For Economic
```{r}
damgeOnEconomic <- tidy[grep("cropdmgvalue|propdmgvalue", tidy[, "damagetype"]),]

ggplot(damgeOnEconomic, 
       aes(x = reorder(event, value), y = value,
           fill = factor(damagetype, labels = c("Property", "Crop")))) + 
        geom_bar(stat = "identity") + 
        labs(title = "Top harmful weather event for economy ",
             x = "Event", y = "Damage on economic value (US dollors)") + 
        guides(fill = guide_legend(title = "Type of damage")) + 
        xlab("Weather Event") + 
        coord_flip()


```

Result: Flash flood are the most severe events with respect to economic damage.